Conversations in time: interactive visualisation to explore structured temporal data

### 2020-08-13 tsibbletalk.Rmd
reworked on categorical linking

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} class with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can contain multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with strongest seasonality, but also a need to querying interesting series on the go. Without interactivity, one needs to first filter the interesting series out from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series to be discovered. Despite that the two plots are static, they can be considered as linked views via the common `key` variables between two tables. This motivates enabling interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

@@

The \pkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk} package, introduces a shared tsibble data structure to enable reactivity for temporal data (i.e. a `tsibble`). The `as_shared_tsibble()` function turns a tsibble to a mutable tsibble `SharedTsibbleData`, as an R6 subclass of `SharedData` from {crosstalk}. A *tsibble* data object holds a set of interrelated series, identified by the "key", in a data frame. In the context of relational databases, the "key" acts like a foreign key in a reduced form of tsibble, while the index and key together operates like a primary key. (tsibble-centered workflow) When the "key" includes more than two variables, there can exist intrinsic structures in all combinations of variables: either nesting or crossing. (Add an example here)

The `spec` argument in the `as_shared_tsibble()` function takes a symbolic formula to allow users to specify structured combinations, using `/` and `*` adopted from Wilkinson notation (10.2307/2346786) for nesting and crossing expressions respectively. By default, the specification assumes crossing structure for all units.

@@

When the key structure involves crossing only, one-to-one linking is the default.

* One-to-many linking: marking a single point of interest highlights all other points that share particular data values (categorical linking)
* Constructing recursive lists. Lists of key values, and json
* Self-linking: a single data source

>>>>

The \CRANpkg{tsibble} package [@wang2020tsibble] introduces a unified temporal data structure, referred to as a `tsibble`, to represent time series and longitudinal data in a tidy format [@wickham2014tidy]. That said, a `tsibble` extends the `data.frame` and \CRANpkg{tibble} class with temporally contextual metadata: `index` and `key`. The `index` declares a data column that holds time-related indices. The `key` identifies a collection of related series or panels observed over the `index`-defined period, which can comprise multiple columns. Below displays the monthly Australian retail trade turnover data (`aus_retail`), available in the \CRANpkg{tsibbledata} package. The `Month` column holds year-months as `index`. The `State` together with `Industry` are the identifiers for these 152 series, highlighted as `key`. Note that the column `Series ID` could be an alternative option for setting up `key`, but `State` and `Industry` are more readable and informative. The `index` and `key` are "sticky" columns to a `tsibble`, forming critical pieces for fluent temporal data analysis later.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with strongest seasonality, but also a need to querying interesting series on the fly. Without interactivity, one needs to first filter the interesting series out from the features table, and join back to the original `tsibble` in order to examine its trend in relation to others. This procedure can soon grow cumbersome if many series to be discovered. Despite that the two plots are static, they can be considered as linked views via the common `key` variables between two tables. This motivates enabling interactivity of `tsibble` and `tsibble`-derived objects for rapid exploratory data analysis.

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk} package, introduces a shared tsibble data structure on top of a `tsibble` to allow for frictionless communication between different plots for temporal data. The `as_shared_tsibble()` function provides an entry point in the flow, turning a `tsibble` to a shared data object (i.e. `SharedTsibbleData` subclassing of `SharedData` from \CRANpkg{crosstalk}) that powers data transmission between plots. The \CRANpkg{tsibbletalk} package aims to streamline interactive graphical analysis with the focus of temporal and structured linking.

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to categorical linking where marking one or more observations in one category will broadcast to all other observations in this category. Given time series plots, click any data point on a line, highlighting the whole line as a result. The `as_shared_tsibble()` uses `tsibble`'s `key` data values to facilitate such linking. The `spec` argument in the `as_shared_tsibble()` function takes a symbolic formula to allow users to specify structured combinations, using `/` and `*` adopted from Wilkinson notation (10.2307/2346786) for nesting and crossing expressions respectively. By default, the specification assumes crossing structure for all units.

@@

* Constructing recursive lists. Lists of key values, and json
* Self-linking: a single data source

### 2020-08-13 tsibbletalk.tex
reworked on categorical linking

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} class with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can contain multiple columns. Below
displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package.
The \texttt{Month} column holds year-months as \texttt{index}. The
\texttt{State} together with \texttt{Industry} are the identifiers for
these 152 series, highlighted as \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
temporal data analysis later.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with
strongest seasonality, but also a need to querying interesting series on
the fly. The two plots are static, but can be considered as linked views
via the common \texttt{key} variables between two tables. This motivates
enabling interactivity of \texttt{tsibble} and \texttt{tsibble}-derived
objects for exploratory data analysis in a lightning manner.

@@

The \pkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk}
package, introduces a shared tsibble data structure to enable reactivity
for temporal data (i.e.~a \texttt{tsibble}). The
\texttt{as\_shared\_tsibble()} function turns a tsibble to a mutable
tsibble \texttt{SharedTsibbleData}, as an R6 subclass of
\texttt{SharedData} from \{crosstalk\}. A \emph{tsibble} data object
holds a set of interrelated series, identified by the ``key'', in a data
frame. In the context of relational databases, the ``key'' acts like a
foreign key in a reduced form of tsibble, while the index and key
together operates like a primary key. (tsibble-centered workflow) When
the ``key'' includes more than two variables, there can exist intrinsic
structures in all combinations of variables: either nesting or crossing.
(Add an example here)

The \texttt{spec} argument in the \texttt{as\_shared\_tsibble()}
function takes a symbolic formula to allow users to specify structured
combinations, using \texttt{/} and \texttt{*} adopted from Wilkinson
notation (10.2307/2346786) for nesting and crossing expressions
respectively. By default, the specification assumes crossing structure
for all units.

@@

When the key structure involves crossing only, one-to-one linking is the
default.

\begin{itemize}
\tightlist
\item
  One-to-many linking: marking a single point of interest highlights all
  other points that share particular data values (categorical linking)
\item
  Constructing recursive lists. Lists of key values, and json
\item
  Self-linking: a single data source
\end{itemize}

>>>>

The \CRANpkg{tsibble} package \citep{wang2020tsibble} introduces a
unified temporal data structure, referred to as a \texttt{tsibble}, to
represent time series and longitudinal data in a tidy format
\citep{wickham2014tidy}. That said, a \texttt{tsibble} extends the
\texttt{data.frame} and \CRANpkg{tibble} class with temporally
contextual metadata: \texttt{index} and \texttt{key}. The \texttt{index}
declares a data column that holds time-related indices. The \texttt{key}
identifies a collection of related series or panels observed over the
\texttt{index}-defined period, which can comprise multiple columns.
Below displays the monthly Australian retail trade turnover data
(\texttt{aus\_retail}), available in the \CRANpkg{tsibbledata} package.
The \texttt{Month} column holds year-months as \texttt{index}. The
\texttt{State} together with \texttt{Industry} are the identifiers for
these 152 series, highlighted as \texttt{key}. Note that the column
\texttt{Series\ ID} could be an alternative option for setting up
\texttt{key}, but \texttt{State} and \texttt{Industry} are more readable
and informative. The \texttt{index} and \texttt{key} are ``sticky''
columns to a \texttt{tsibble}, forming critical pieces for fluent
temporal data analysis later.

@@

Figure \ref{fig:highlight-retail} highlights not only a series with
strongest seasonality, but also a need to querying interesting series on
the fly. Without interactivity, one needs to first filter the
interesting series out from the features table, and join back to the
original \texttt{tsibble} in order to examine its trend in relation to
others. This procedure can soon grow cumbersome if many series to be
discovered. Despite that the two plots are static, they can be
considered as linked views via the common \texttt{key} variables between
two tables. This motivates enabling interactivity of \texttt{tsibble}
and \texttt{tsibble}-derived objects for rapid exploratory data
analysis.

@@

The \CRANpkg{tsibbletalk} package, inspired by the \CRANpkg{crosstalk}
package, introduces a shared tsibble data structure on top of a
\texttt{tsibble} to allow for frictionless communication between
different plots for temporal data. The \texttt{as\_shared\_tsibble()}
function provides an entry point in the flow, turning a \texttt{tsibble}
to a shared data object (i.e.~\texttt{SharedTsibbleData} subclassing of
\texttt{SharedData} from \CRANpkg{crosstalk}) that powers data
transmission between plots. The \CRANpkg{tsibbletalk} package aims to
streamline interactive graphical analysis with the focus of temporal and
structured linking.

As opposed to one-to-one linking, \CRANpkg{tsibbletalk} defaults to
categorical linking where marking one or more observations in one
category will broadcast to all other observations in this category.
Given time series plots, click any data point on a line, highlighting
the whole line as a result. The \texttt{as\_shared\_tsibble()} uses
\texttt{tsibble}'s \texttt{key} data values to facilitate such linking.
The \texttt{spec} argument in the \texttt{as\_shared\_tsibble()}
function takes a symbolic formula to allow users to specify structured
combinations, using \texttt{/} and \texttt{*} adopted from Wilkinson
notation (10.2307/2346786) for nesting and crossing expressions
respectively. By default, the specification assumes crossing structure
for all units.

@@

\begin{itemize}
\tightlist
\item
  Constructing recursive lists. Lists of key values, and json
\item
  Self-linking: a single data source
\end{itemize}
